---
title: "Creating a Synthetic Control and Preparing Data for SCITS Analysis"
author: "Kate Vinita Fitch"
output:
  html_document: default
---

# Package and dataset import

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(Synth)

dat <- readRDS("./dat_deidentified.rds")


```

First let's view the data so that we can understand how it's structured. Fortunately we've done the cleaning for you!

```{r, echo = FALSE}

as_tibble(dat)

```

### As you look through the data, here are some things to observe:

- There is one row for study time point, from 96 months prior to intervention to 18 months after intervention
- We have one intervention county called "INTERVENTION" with id of 1
- We have 12 control counties with ids of 2-13
- Group = 1 in the intervention county and 0 in the control counties
- Policy = 1 in the intervention county after the time of intervention
- We have several predictor variables related to county demographics and other features
- We have one outcome variable called "outcome"

### Let's build our synthetic controls!

We will be telling the algorithm to build a synthetic control based on several predictor variables that maximizes the trend fit prior to the intervention time. 

```{r, results = FALSE, message = FALSE, warning = FALSE}

dataprep.out <- dataprep(foo = dat, # data set name
                         predictors = c("pop"), # select an initial predictor
                         predictors.op = "mean", # we are modeling based on mean value in each county
                         time.predictors.prior = c(-96:-1), # note the study time prior to intervention
                         special.predictors = list( # add more predictors
                           
                            #study time prior to intervention, every 12 months
                           list("pct_male", seq(-96,-1,12), "mean"),
                           list("pct_aian", seq(-96,-1,12), "mean"),
                           list("pct_asian", seq(-96,-1,12), "mean"),
                           list("pct_black", seq(-96,-1,12), "mean"),
                           list("pct_other", seq(-96,-1,12), "mean"),
                           list("male_unemp", seq(-96,-1,12), "mean"),
                           list("total_unemp", seq(-96,-1,12), "mean"),
                           list("fm_unemp_ratio", seq(-96,-1,12), "mean"),
                           list("bachelors", seq(-96,-1,12), "mean"),
                           list("median_household_income", seq(-96,-1,12), "mean"),
                           list("officers_per_100k", seq(-96,-1,12), "mean"),
                           
                            # Rural occurs only once during the census year, 
                           # so we will just average based on one month in that period
                      
                           list("rural", -76, "mean")
                         ),
                         
                         dependent = "outcome", # name the outcome variable
                         unit.variable = "id", # county identifier
                         unit.names.variable = "county", # county name
                         time.variable = "study_time", # time variable
                         treatment.identifier = 1, # id 1 is the intervention county
                         controls.identifier = c(2:13), # the remaining ids are control counties
                         time.optimize.ssr = c(-96:-1), # optimize fit from study times -96 to -1
                         time.plot = c(-96:18)) # produce predictive values up to time 18

# Outputting the synthetic control algorithm results
synth.out <- synth(dataprep.out)

# Outputting the weights of the 12 control counties and attaching to it the names of those counties
weights <- round(synth.out$solution.w,4)
weights <- cbind(county_id = rownames(weights), weights)

# Outputting the case (intervention) and synthetic control outcome values
path.case <- dataprep.out$Y1plot
path.synth <- dataprep.out$Y0plot %*% synth.out$solution.w

```

Below we can see how the donor counties were weighted in building the synthetic control county.

```{r, echo = FALSE}

as_tibble(weights)

```

## Constructing dataset for use in time series analysis

Now that we've constructed the synthetic control, we need to construct a dataset that includes the outcome variable at each time point for both the intervention and synthetic control counties.

First, restrict the analysis dataset to only the county, study time, and outcome.

```{r}

outcome <- dat %>%
  dplyr::select(county, study_time, outcome)

as_tibble(outcome)

```

Next, Widen the dataset to have only one row per time with each county's outcome in a column.

```{r}

outcome_wide <- pivot_wider(outcome, names_from = county, values_from = outcome)

as_tibble(outcome_wide)

```
Next, we'll append the outcome variables for the synthetic control (path.synth) and intervention (path.case) counties from the Synth package. Importantly, the path.case column should match the intervention county's column and that's a good visual check to perform here.

```{r}

# Make the intervention and synthetic control outcome variable list into a dataframe with the column header indicating its identity

path.case <- as.data.frame(path.case)
names(path.case)[1] <- "path.case"
path.synth <- as.data.frame(path.synth)
names(path.synth)[1] <- "path.synth"

# Merge the intervention and synthetic control outcome data to the wide outcome dataset
dat_synth <- cbind(outcome_wide, path.case, path.synth)

# Remove the (non-synthetic) control county columns
dat_synth <- dat_synth %>%
  select(study_time, INTERVENTION, path.case, path.synth)
  

```

```{r, echo = FALSE}

as_tibble(dat_synth)

```

Now we'll need to shift back from wide to long with separate entries at each time point for the intervention and synthetic control counties. First we'll create a new column called "group" and stack the outcome variable for path.case and path.synth instead of having a separate column for each.

```{r}

dat_synth_long <- gather(dat_synth, group, outcome, c(path.case, path.synth),
                       factor_key = TRUE)

```

```{r, echo = FALSE}

as_tibble(dat_synth_long)

```

Now we will change the group column to make it an indicator variable with 1 for the intervention county (path.case) and 0 for the synthetic control county (path.synth). We will also drop the INTERVENTION column, as we no longer need it.

```{r}

dat_synth_long <- dat_synth_long %>%
  mutate(group = ifelse(group == "path.case", 1, 0)) %>%
  select(-INTERVENTION)

```

```{r, echo = FALSE}

as_tibble(dat_synth_long)

```
## Creating variables for time series analysis

We need to create a number of additional variables for the time series analysis done in SAS, and can easily do so with the dplyr::mutate function.

```{r}

scits_dat <- dat_synth_long %>%
  mutate(
    # Policy = 1 if study_time is > 0 
    policy = ifelse(study_time > 0, 1, 0),
    
    # Posttime is equal to study time only after intervention, else 0
    posttime = ifelse(study_time > 0, study_time, 0),
    
    # Group time is equal to the study time in the intervention group, else 0
    group_time = ifelse(group == 1, study_time, 0),
    
    # Group policy is equal to policy in the intervention group, else 0
    group_policy = ifelse(group == 1, policy, 0),
    
    # Group pposttime is equal to the posttime in the intervention group, else 0
    group_posttime = ifelse(group ==1, posttime, 0)) %>%
  
  # Organize the columns for easy viewing
  select(study_time, outcome, group, policy, posttime, group_time, group_policy, group_posttime)

```

```{r, echo = FALSE}

as_tibble(scits_dat)

```
### Now export the data into a .csv file for easy use in SAS!
```{r export}


write.csv(scits_dat, "./dummy_scits.csv")

```
